The Rockerverse: Packages and Applications for Containerisation with R

### 2020-03-02 manuscript.Rmd
Small edits, PS to authors

@@

The R community continues to grow.
This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Also though, coinciding with the rise of R has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
As such, this article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and environment with other containerisation solutions.
Our main contributions is a coherent picture of the current status of using containers in, with, and for R.

The article continues with a brief introduction of containerisation basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specificly for interacting with Docker, next the second-level packages that use containers indirectly or only for specific features, and finally some complex use cases that leverage containers.
We conclude with by reflecting on the landscape of packages and applications and point out future directions of development.

@@

Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, Docker provides an important scalability opportunity.
Beyond this though, Docker has improved this compartmentalisation by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualisation.
This containerisation, also called operating-system-level virtualisation \citep{wikipedia_contributors_os-level_2020}, 
makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018} because, while a container typically runs one process, virtualisation may run whole operating systems at a larger footprint.
This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
While Docker makes use of Linux kernel features, it has become important enough that some required aspects of running Docker have been added to other operating systems so that those systems can more efficiently support Docker \citep{microsoft_linux_2019}.
The success of Docker has even led to standardisation and industry collaboration \citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since it began in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for easier maintainability (i.e., smaller `Dockerfile`s), better composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, which provides funding for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images, and launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
As such, the choice of which package is most useful depends on the use case at hand as well as on the user's level of expertise.

@@

functionalities <- tribble(
  ~id, ~Functionality,
  "dockerfile", "Generate a Dockerfile",
  "build", "Build an image",
  "execute", "Execute a container locally or remotely",
  "cloud", "Deploy or manage an instances in the cloud",
  "interact", "Interact with an instance (e.g., file transfer)",
  "images", "Manage storage of images",
  "othercontainers", "Supports Docker and Singularity",
  "api", "Direct access to Docker API instead of using the CLI",
  "install", "Installing Docker software"
)

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The output of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
```{r harbor, eval=FALSE, include=FALSE}
remotes::install_github("wch/harbor")
harbor::docker_run(image = "hello-world")
```

@@

\label{datascience}
**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the data science community led to Docker being used widely.
<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
The remaineder of this section presents exemple Docker images and image stacks featuring R intended for data science.

@@

System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R version and the the R packages.
R versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions and for historic exploration of R.
With a standalone NodeJS app or [online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} and **\CRANpkg{renv}**, which can naturally be applied within images and containers.
For example, \CRANpkg{renv} ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
While \CRANpkg{renv} does not cover system dependencies, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work indepenently of physical computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.} and to assert a specific degree of confidence in the stability of a developed software \citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.

@@

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
Purpose-built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
```{r geospatial_deps, eval=FALSE, include=FALSE}
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
```

@@

Fourth, containers are useful for **testing** R code during development.
To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally; this can be a challenge for some users.
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one \citep{r-hub_docs_2019}.
R-hub makes it easy to ensure that no errors occur, but fixing errors still often warrants a local setup, e.g., using the image `rocker/r-devel`, as is testing packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but they lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems with which they are not familiar.
The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied with a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
This approach clearly seperates test environment, test code, and current working copy of the code.
\label{rselenium} Another use case where a container can help to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but it requires users to manually manage the containers.

@@

**\CRANpkg{batchtools}** provides a parallel implementation of the Map function for various schedulers \citep{Lang2017batchtools}.

@@

The cloud is the natural environment of containers, and, subsequently, containers are the go-to mechanism to deploy R server applications.
More and more continuous integration (CI) and continuous delivery (CD) services also use containers, opening up new options for use.
The controlled nature of containers, i.e., the possibility to abstract internal software environment from a minimal dependency outside of the container is crucial, for example to match test or build environments with production environments or transfer runnable entities to as-a-service infrastructures.

@@

R has been historically viewed as a tool for analysis and scientific research, but not for creating software that corporations can rely on for production services.
However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
A number of packages and projects have enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
The latter two also provide documentation for deployment with Docker or ready-to-use images with automated builds\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
These software allow other (remote) processes and programming languages to interact with R and to expose R-based function in a service architecture with HTTP APIs.
APIs based on these packages can be deployed with scalability and high availability using containers.
This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first-class member of a software engineering technical stack.

@@

The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
For example, one model quickly identifies whether a message is from a customer \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
The models are convolutional neural networks that use the \CRANpkg{keras} package and run in a Rocker Docker image.
The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
The AI @ T-Mobile team found that the models performed well, and today these models power real-time services that are called over a million times a day.

@@

\label{binder}
The [**Binder**](https://mybinder.readthedocs.io/en/latest/) project, maintained by the team behind Jupyter, makes it possible for users to **create and share computing environments** with others \citep{jupyter_binder_2018}.
A _BinderHub_ allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any workflows contained within.
From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
Binder's core internal tool is `repo2docker`.
It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config\_files.html}{https://repo2docker.readthedocs.io/en/latest/config\_files.html}. For R, the }.
In the most powerful case, `repo2docker` builds a given `Dockerfile`.
While this approach works well for most run-of-the-mill Python projects, it is not so seamless for R projects.
This is partly because `repo2docker` does not support arbitrary base images due to the complex auto-generation of the `Dockerfile` instructions.
Two approaches make using Binder easier for R users.
First, **\pkg{holepunch}** ([https://github.com/karthik/holepunch](https://github.com/karthik/holepunch)) is an R package that was designed to make sharing work environments accessible to novice R users based on Binder.
For any R projects that use the Tidyverse suite \citep{wickham_welcome_2019}, the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.
\pkg{holepunch} removes some of these limitations by leveraging Rocker images that contain the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images.
It short cicuits the configuration file parsing in `repo2docker` and starts with the Binder/Tidyverse base images, which eliminates a large part of the build time and, in most cases, results in a Binder instance launching within a minute.

>>>>

  - name: Péter Sólymos
    # https://github.com/psolymos
    affiliation: Analythium Solutions
      - \#258 150 Chippewa Road 
      - Sherwood Park, AB, T8A 6A2, Canada
      - \orcid{0000-0001-7337-1740}
    email: peter@analythium.io

@@

The R community continues to grow.
This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Also though, coinciding with the rise of R has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides Docker images with R (see the next section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
As such, this article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and environment with other containerisation solutions.
Our main contributions is a coherent picture of the current status of using containers in, with, and for R.

The article continues with a brief introduction of containerisation basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specifically for interacting with Docker, next the second-level packages that use containers indirectly or only for specific features, and finally some complex use cases that leverage containers.
We conclude with by reflecting on the landscape of packages and applications and point out future directions of development.

@@

Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operating systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, Docker provides an important scalability opportunity.
Beyond this though, Docker has improved this compartmentalisation by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualisation.
This containerisation, also called operating-system-level virtualisation \citep{wikipedia_contributors_os-level_2020}, 
makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018} because, while a container typically runs one process, virtualisation may run whole operating systems at a larger footprint.
This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
While Docker makes use of Linux kernel features, it has become important enough that some required aspects of running Docker have been added to other operating systems so that those systems can more efficiently support Docker \citep{microsoft_linux_2019}.
The success of Docker has even led to standardisation and industry collaboration \citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since it began in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for easier maintainability (i.e., smaller `Dockerfile`s), better composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, which provides funding for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images, and launching containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or in the cloud.
As such, the choice of which package is most useful depends on the use case at hand as well as on the user's level of expertise.

@@

functionalities <- tribble(
  ~id, ~Functionality,
  "dockerfile", "Generate a Dockerfile",
  "build", "Build an image",
  "execute", "Execute a container locally or remotely",
  "cloud", "Deploy or manage instances in the cloud",
  "interact", "Interact with an instance (e.g., file transfer)",
  "images", "Manage storage of images",
  "othercontainers", "Supports Docker and Singularity",
  "api", "Direct access to Docker API instead of using the CLI",
  "install", "Installing Docker software"
)

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor)) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The outputs of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
```{r harbor, eval=FALSE, include=FALSE}
remotes::install_github("wch/harbor")
harbor::docker_run(image = "hello-world")
```

@@

\label{datascience}
**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the data science community has led to Docker being used widely.
<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
The remaineder of this section presents exemple Docker images and image stacks featuring R intended for data science.

@@

System dependencies are the domain of Docker, but for a full description of the computing environment one must also manage the R version and the the R packages.
R versions are available via the versioned Rocker image stack \citep{RJ-2017-065}.
[r-online](https://github.com/ColinFay/ronline) leverages these images and provides an app for helping users to detect breaking changes between different R versions and for historic exploration of R.
With a standalone NodeJS app or [r-online](https://srv.colinfay.me/r-online), the user can compare a piece of code run in two separate versions of R.
Internally, r-online opens one or two Docker instances with the given version of R based on Rocker images, executes a given piece of code, and returns the result to the user.
Regarding R package management, this can be achieved with MRAN, or with packages such as \CRANpkg{checkpoint} and **\CRANpkg{renv}**, which can naturally be applied within images and containers.
For example, \CRANpkg{renv} ([https://rstudio.github.io/renv/](https://rstudio.github.io/renv/)) helps users to manage the state of the R library in a reproducible way, further providing isolation and portability \citep{renv2019}.
While \CRANpkg{renv} does not cover system dependencies, the \CRANpkg{renv}-based environment can be transferred into a container either by restoring the environment based on the main configuration file `renv.lock` or by storing the \CRANpkg{renv}-cache on the host and not in the container \citep{ushey_using_2019}.
With both the system dependencies and R packages consciously managed in a Docker image, users can start using containers as the _only_ environment for their workflows, which allows them to work indepenently of physical computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.} and to assert a specific degree of confidence in the stability of a developed software \citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.

@@

Second, the strong integration of **system libraries in core packages** in the [R-spatial community](https://www.r-spatial.org/) makes containers essential for stable and proactive development of common classes for geospatial data modelling and analysis. For example, GDAL \citep{gdal_2019} is a crucial library in the geospatial domain.
GDAL is a system dependency allowing R packages such as \CRANpkg{sf}, which provides the core data model for geospatial vector data, or \CRANpkg{rgdal}, to accomodate users to be able to read and write hundreds of different spatial raster and vector formats.
\CRANpkg{sf} and \CRANpkg{rgdal} have hundreds of indirect reverse imports and dependencies and, therefore, the maintainers spend a lot of effort trying not to break them.
Purpose-built Docker images are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.
```{r geospatial_deps, eval=FALSE, include=FALSE}
length(tools::dependsOnPkgs(c("rgdal", "sf"), dependencies = c("most"), recursive = TRUE))
```

@@

Fourth, containers are useful for **testing** R code during development.
To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally; this can be a challenge for some users.
The **R-hub** project provides _"a collection of services to help R package development"_, with the package builder as the most prominent one \citep{r-hub_docs_2019}.
R-hub makes it easy to ensure that no errors occur, but fixing errors still often warrants a local setup, e.g., using the image `rocker/r-devel`, as is testing packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but they lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems with which they are not familiar.
The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied by a special launch script so the tested source code is not stored in the image; instead, the currently checked in version from a local Git repository is cloned into the container at runtime.
This approach clearly seperates test environment, test code, and current working copy of the code.
\label{rselenium} Another use case where a container can help to standardise tests across operating systems is detailed the vignettes of the package \CRANpkg{RSelenium} \citep{rselenium_2019}.
The package recommends Docker for running the [Selenium](https://selenium.dev/) Server application needed to execute test suites on browser-based user interfaces and webpages, but it requires users to manually manage the containers.

@@

**\CRANpkg{batchtools}** provides a parallel implementation of the `Map` function for various schedulers \citep{Lang2017batchtools}.

@@

The cloud is the natural environment for containers, and, therefore, containers are the go-to mechanism for deploying R server applications.
More and more continuous integration (CI) and continuous delivery (CD) services also use containers, opening up new options for use.
The controlled nature of containers, i.e., the possibility to abstract internal software environment from a minimal dependency outside of the container is crucial, for example to match test or build environments with production environments or transfer runnable entities to as-a-service infrastructures.

@@

R has been historically viewed as a tool for analysis and scientific research, but not for creating software that corporations can rely on for production services.
However, thanks to advancements in R running as a web service, along with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
A number of packages and projects have enabled R to respond to client reqests over TCP/IP and local socket servers, such as \CRANpkg{Rserve}, \CRANpkg{svSocket}, [rApache](http://www.rapache.net) and more recently \CRANpkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) and \pkg{RestRserve} ([http://restrserve.org](http://restrserve.org)), which even processes incoming requests in parallel with forked processes using \CRANpkg{Rserve}.
The latter two also provide documentation for deployment with Docker or ready-to-use images with automated builds\footnote{See \href{https://www.rplumber.io/docs/hosting.html\#docker}{https://www.rplumber.io/docs/hosting.html\#docker}, \href{https://hub.docker.com/r/trestletech/plumber/}{https://hub.docker.com/r/trestletech/plumber/} and \href{https://hub.docker.com/r/rexyai/restrserve/}{https://hub.docker.com/r/rexyai/restrserve/}.}.
These software allow other (remote) processes and programming languages to interact with R and to expose R-based function in a service architecture with HTTP APIs.
APIs based on these packages can be deployed with scalability and high availability using containers.
This pattern of deploying code matches those used by software engineering services created in more established languages in the enterprise domain, such as Java or Python, and R can be used alongside those langages as a first-class member of a software engineering technical stack.

@@

The AI @ T-Mobile team created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers \citep{t-mobile_enterprise_2018}.
For example, one model quickly identifies whether a message is from a customer \citep[see \CRANpkg{Shiny}-based \href{https://secure.message.t-mobile.com/v1/shiny/is-customer/app/}{demo} further described by ][]{nolis_small_2019}, and others tell which customers are likely to make a repeat purchase.
If a data scientist creates a machine learning model and exposes it through a \CRANpkg{plumber} API, then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
The models are convolutional neural networks that use the \CRANpkg{keras} package and run in a Rocker Docker container.
The corresponding `Dockerfile`s are published [on GitHub](https://github.com/tmobile/r-tensorflow-api).
Since the models power tools for agents and customers, they need to have extremely high uptime and reliability.
The AI @ T-Mobile team found that the models performed well, and today these models power real-time services that are called over a million times a day.

@@

\label{binder}
The [**Binder**](https://mybinder.readthedocs.io/en/latest/) project, maintained by the team behind Jupyter, makes it possible for users to **create and share computing environments** with others \citep{jupyter_binder_2018}.
A _BinderHub_ allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any workflows contained within.
From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.
To set up Binder for a project, a user typically starts at an instance of a BinderHub and passes the location of a repository with a workspace, e.g., a hosted Git repository, or a data repository like Zenodo.
Binder's core internal tool is `repo2docker`.
It deterministically builds a Docker image by parsing the contents of a repository, e.g., project dependency configurations or simple configuration files\footnote{See supported file types at \href{https://repo2docker.readthedocs.io/en/latest/config\_files.html}{https://repo2docker.readthedocs.io/en/latest/config\_files.html}. For R, the }.
In the most powerful case, `repo2docker` builds a given `Dockerfile`.
While this approach works well for most run-of-the-mill Python projects, it is not so seamless for R projects.
This is partly because `repo2docker` does not support arbitrary base images due to the complex auto-generation of the `Dockerfile` instructions.
